---
title: "class 5: Data Viz with ggplot"
author: "Sahar Kamali (PID: A19156624)"
format: pdf
toc: true
---

## Background 

There are lot's of ways to make plots in R. These include so-called "base R" (like the `plot()`) and add on packages like **ggplot**, 

Let's make the same plot with these two graphics systems. We can use the inbuilt `car` dataset: 

```{r}
head( cars )
```

With "base R" we can simply: 

```{r}
plot (cars)
```

Now let's try ggplot. First i need to install the package using `install.packages("ggplot2")`

> **N.B.** We never run on `install.packages("ggplot2")`in a code chunk otherwise we will re-install needlesly every time we render our document. 


Every time we want to use an add-on package we need to load it up with a call to `library()`

```{r}
library(ggplot2)
ggplot(cars)
```


Every ggplot needs at least 3 things: 

1. The **data** i.e. stuff to plot as a data.frame 
2. The **aes** or aesthetics that map the data to the plot 
3. The **geom_** or geometry i.e. the plot type such as points,lines etc. 

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +
geom_point() + 
  geom_line()
```


```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +
geom_point() + 
  geom_smooth()
```


```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point() + 
  geom_smooth(method= "lm", se= FALSE) + labs(x="speed (MPH)", y= "Distance(ft)", title= "Stopping Distance of old Cars") 
```

###Gene Expression Plot 

Read some data on the effects of BLP-1 inhibitor (drug) on gene expression values: 

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

Version 1 plot - start simple by getting some ink on the page. 

```{r}
ggplot(genes) + 
  aes(x=Condition1, y=Condition2,) + 
  geom_point(col="blue", alpha=0.2)
```

Let's color by `State` up, down or not change. 

```{r}
table( genes$State ) 
```


```{r}
ggplot(genes) + 
  aes(x=Condition1, y=Condition2, col=State) + 
  geom_point() + 
  scale_color_manual(values= c("purple", 
                               "grey",
                               "orange"))+ 
labs(x= "Control (no drug", 
     y= "Drug", 
     title= "Expression chnages with GLP-1 drug")
theme_classic()
```


## Going further with gapmider 

Here we explore the famous `gapminder` data sets with some custom plots. 


```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```

>Q. How many rows does this data set have? 

```{r}
nrow(gapminder)
```


> How many different continents are in this date set? 

```{r}
table(gapminder$continent)
```


version 1 plot gdppercap vs LifeExp for all rows

```{r}
ggplot (gapminder) + 
  aes(gdpPercap, lifeExp, col= continent) + 
  geom_point() 
```




I want to see a plot for each continents - in ggplot lingo this is called "faceting" 

```{r}
ggplot (gapminder) + 
  aes(gdpPercap, lifeExp, col= continent) + 
  geom_point() + 
  facet_wrap(~ continent)
```

## First look at the dplyr package 

Another add-on package with a function called `filter()` that we want to use. 

```{r}
library(dplyr)
```


```{r}
input <-filter(gapminder, year==2007 | year==1977)
library(ggplot2)
ggplot (input) + 
  aes(gdpPercap, lifeExp, col= continent) + 
  geom_point() + 
  facet_wrap(~ year)
```








