---
title: "Class8 Mini Project"
author: Sahar kamali (PID:A19156624)
format: pdf
toc: true
---

## Preparing Data/ Exploratory data analysis

```{r}
fna.data <- "WisconsinCancer .csv"

# Read the CSV file into R
wisc.df <- read.csv(fna.data, row.names = 1)
```



```{r}
# Pull out diagnosis column (M or B)
diagnosis <- wisc.df$diagnosis

# Create a features-only data frame (remove diagnosis column)
wisc.data <- wisc.df[, -1]
```

```{r}
table(diagnosis)
dim(wisc.data)
```


```{r}
# Perform PCA on the scaled data
wisc.pr <- prcomp(wisc.data, center = FALSE, scale. = TRUE)
```



```{r}
summary(wisc.pr)
```

Questions: 

```{r}
# Q1: number of observations
nrow(wisc.data)

# Q2: number of malignant observations
table(diagnosis)["M"]

# Q3: number of variables ending in "_mean"
length(grep("_mean$", colnames(wisc.data)))
```
###Q1.  How many observations are in this dataset?
10 observations. 

###Q2.  How many of the observations have a malignant diagnosis?
212 malignant diagnosis. 

###Q3. How many variables/features in the data are suffixed with _mean?
10 variables. 



## Performing PCA and results 

```{r}
attributes(wisc.pr)
```
```{r}
head(wisc.pr$x)
```

Questions: 
###Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?
44.3% of the total variance or 0.4427

###Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?
3 principal components.
Cumulative variance after:
PC1 = 44.27%
PC2 = 63.24%
PC3 = 72.64%

###Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?
7 principal components.
Cumulative variance after:
PC6 = 88.76%
PC7 = 91.01%

###Q7.What stands out to you about this plot? Is it easy or difficult to understand? Why?
The biplot is difficult to interpret because it is very cluttered and there are many overlapping points.

## Variance explained 

```{r}
library(ggplot2)

ggplot(as.data.frame(wisc.pr$x)) +
  aes(PC1, PC2, col = diagnosis) +
  geom_point()
```

```{r}
library(ggplot2)

ggplot(as.data.frame(wisc.pr$x)) +
  aes(PC1, PC3, col = diagnosis) +
  geom_point()
```

###Q8. Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?
In the PC1 vs PC3 plot, the two groups still separate mostly along PC1. PC3 doesn’t add much extra separation the benign and malignant points overlap a lot vertically so the main pattern is basically the same as the PC1 vs PC2 plot.PC1 is doing most of the work.


```{r}
# Calculate variance of each component
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

```{r}
# Variance explained by each principal component: pve
pve <- pr.var / sum(pr.var)

# Plot variance explained for each principal component
plot(c(1, pve),
     xlab = "Principal Component",
     ylab = "Proportion of Variance Explained",
     ylim = c(0, 1),
     type = "o")
```


```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve,
        ylab = "Percent of Variance Explained",
        names.arg = paste0("PC", 1:length(pve)),
        las = 2,
        axes = FALSE)

axis(2, at = pve, labels = round(pve, 2) * 100)
```


```{r}
# OPTIONAL: PCA variance explained plot using factoextra
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```


```{r}
# Q9: PC1 loading for concave.points_mean
wisc.pr$rotation["concave.points_mean", 1]
```

###Q9. Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?
For PC1, the loading of concave.points_mean is approximately −0.17, meaning this feature contributes to PC1 but is not the strongest contributor. Other features have larger loadings and therefore have a bigger influence on PC1.

Hierarchical clustering

```{r}
# Check which features have the largest absolute PC1 loadings
pc1_loadings <- wisc.pr$rotation[, 1]

# Sort by absolute contribution
sort(abs(pc1_loadings), decreasing = TRUE)[1:10]
```

## Hierarchical clustering and combining methods

```{r}
# Scale the wisc.data using the scale() function
data.scaled <- scale(wisc.data)
```

```{r}
# Calculate Euclidean distances between observations
data.dist <- dist(data.scaled)
```

```{r}
# Perform hierarchical clustering using complete linkage
wisc.hclust <- hclust(data.dist, method = "complete")
```

```{r}
wisc.hclust.clusters<- cutree(wisc.hclust ,k=4) 
```



```{r}
# Plot the hierarchical clustering dendrogram
plot(wisc.hclust, labels = FALSE, main = "Hierarchical Clustering Dendrogram")

# Add a horizontal line to indicate the cut height
abline(h = 20, col = "red", lty = 2)
```


```{r}
# Plot the dendrogram
plot(wisc.hclust, labels = FALSE, main = "Hierarchical Clustering Dendrogram")

# Add a horizontal cut line (chosen to give 4 clusters)
abline(h = 15, col = "red", lty = 2)
```


```{r}
# Cut the tree to get 4 clusters
pc.dist<-dist(wisc.pr$x[,1:3])
wisc.pr.hclust<-hclust(pc.dist, method="ward.D2")
plot(wisc.pr.hclust)
```


###Q10. Using the plot() and abline() functions, what is the height at which theclustering model has 4 clusters?
From the dendrogram, the tree is cut at a height of approximately 15 to produce 4 clusters, as indicated by the horizontal red line where four main branches remain separated.

##Q11. OPTIONAL: Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 6? How do you judge the quality of your result in each case?
Cluster 1 largely corresponds to malignant (M) samples, while cluster 3 largely corresponds to benign (B) samples, indicating that the hierarchical clustering separates the two diagnoses reasonably well.

###Q12.Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.
Among the linkage methods tested, ward.D2 gives the best results because it produces tighter, more distinct clusters with better separation between benign and malignant samples compared to single or complete linkage. 


```{r}
grps<-cutree(wisc.pr.hclust, k=2)
table(grps)
```

```{r}
table(grps,diagnosis)
```

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k = 4)
```

```{r}
# Compare cluster assignments to diagnosis
table(wisc.hclust.clusters, diagnosis)
```


```{r}
# Compare different linkage methods


for (m in 2:6) {
  cat("\n====================\n")
  cat("Linkage method:", m, "\n")
  
  hc <- hclust(data.dist, method = "ward.D2")
  clusters <- cutree(hc, k = m)
  
  print(table(clusters, diagnosis))
}
```



```{r}
 hc <- hclust(data.dist, method = "single")
  clusters <- cutree(hc, k = 2)
  
  print(table(clusters, diagnosis))
```

```{r}
 hc <- hclust(data.dist, method = "complete")
  clusters <- cutree(hc, k = 2)
  
  print(table(clusters, diagnosis))
```
```{r}
 hc <- hclust(data.dist, method = "average")
  clusters <- cutree(hc, k = 2)
  
  print(table(clusters, diagnosis))
```


```{r}
## Use the distance along the first 7 PCs for clustering i.e. wisc.pr$x[, 1:7]
wisc.pr.hclust <- hclust(dist(wisc.pr$x[, 1:3]), method = "ward.D2")
```


```{r}
## Cut this hierarchical clustering model into 2 clusters
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k = 2)

## Check cluster sizes
table(wisc.pr.hclust.clusters)
```

```{r}
## Compare PCA-based clusters to actual diagnoses
table(wisc.pr.hclust.clusters, diagnosis)
```

```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```

## Sensitivity/Specificity

I can run `table()` with both my clustering `grps` and the expert `diagnosis` 

```{r}
table(grps, diagnosis)
```

our cluster "1" has 179 "M" diagnosis 
our cluster "2" has 333 "B" diagnosis 

179 TP 
24 FP 
333 TN 
33 FN 

Sensitivity: TP/(TP+FN)
```{r}
179/(179+33)
```

Specificity: TN/(TN+FP)
```{r}
333/(333+24)
```

###Q13. How well does the newly created hclust model with two clusters separate out the two “M” and “B” diagnoses?
The two-cluster model separates the diagnoses fairly well. One cluster contains mostly malignant samples, while the other contains mostly benign samples, with only a small number of samples misclassified in each group. Overall, this shows strong separation between M and B using two clusters.

###Q14. How well do the hierarchical clustering models you created in the previous sections (i.e. without first doing PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.hclust.clusters and wisc.pr.hclust.clusters) with the vector containing the actual diagnoses. 
The hierarchical clustering model with PCA performs better than the model without PCA. Using PCA first leads to cleaner separation between malignant and benign samples, while clustering on the original scaled data produces more mixed clusters. Overall, PCA improves the clustering results.



```{r}
table(wisc.hclust.clusters, diagnosis)

```



## Prediction/ Summary

```{r}
#url <- "new_samples.csv"
wisc.pr<-prcomp(wisc.data, scale=T)
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- stats::predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

###Q16. Which of these new patients should we prioritize for follow up based on your results?
Patient 2 should be prioritized for follow up because it falls closer to the cluster dominated by malignant samples in the PCA plot. Patient 1 is located closer to the benign cluster and appears less concerning by comparison.

