---
title: "Pertussis and the CMI-PB project"
author: "Sahar Kamali (PID:A19156624)"
format: pdf
---

##Background 
Pertussis (Whooping cough) is a common lung infection caused by the bacteria B. Pertussis. It can infect everyone but is most deadly for infants (under 1 year). 

# CDC tracking data 

the CDC tracks the number of pertussis cases: 

```{r, echo=FALSE}
cdc <- data.frame(
                                 year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L,2022L,2023L,2024,2025),
         cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116,3044,7063,22538,21996)
       )
```




>Q1. I want a plot of year vs cases

```{r}
library(ggplot2)

ggplot(cdc) +
  aes(year, cases) +
geom_point() +
geom_line()
```


>Q2. Add annotation line for the major milestone of wP vaccination roll-out (1946) and switch to the aP vaccine9 (1996). >Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
library(ggplot2)

ggplot(cdc) +
  aes(year, cases) +
geom_point() +
geom_line() +
geom_vline(xintercept = 1946, col="blue", lty=2) + 
geom_vline(xintercept = 1996, col="red", lty=2) +
geom_vline(xintercept = 2020, col="gray", lty=2)
```



>Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the aP vaccine was introduced, pertussis cases started increasing again. This could be because immunity from the aP vaccine may wear off faster, better PCR testing detects more cases, and the bacteria may have evolved to escape vaccine immunity.

##Exploring CMI-PB data
the CMI- PB project < https://www.cmi-pb.org/ > The mission of CMI-PB is to provide the scientific community with a comprehensive, high-quality and freely accessible resource of Pertussis booster vaccination.

Basically, make available a large datasets on the immune response to Pertussis. They use a "booster" vaccination as a proxy for Pertussis infection. 

They make their data available as JSON format API. We can read this into R with the ` rea_JSON()` function from the **jsonlite** package: 

```{r}
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/v5_1/subject", simplifyVector = TRUE)

head(subject)

```



>Q4. How many aP and wP individuals are there in this `subject` table? 

```{r}
table(subject$infancy_vac)
```


>Q5. How many Male and Female subjects/patients are in the dataset?
How many male/female are there?


```{r}
table(subject$biological_sex)
```

>Q. What is the break down of `biological_sex` gender and `race`?

>Q. Is this representative of the US population? 
>Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```

We can read more table from the CMI-PB database 

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/v5_1/specimen", simplifyVector = TRUE )
ab_titer <- read_json("https://www.cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector = TRUE)

head(specimen)
```


```{r}
head(ab_titer)
```


To make sense of all this data we need to "join" (a.k.a "merge" or "link") all these tables together. Only then will you know that a given Ab measurement (from the `ab_titer` table) was collected on a certain data (from the `specimen` table) from a certain wP or aP subject (from the `subject` table).  

We can use **dplyer** as the `*_join()` family of functions to do this. 

```{r}
library(dplyr)
meta <- inner_join(subject, specimen)
head(meta)
```

Let's do one more `inner_join()` to join the `ab_titer` with all our `meta` data. 

```{r}
abdata <- inner_join(ab_titer, meta)
head(abdata)
```

>Q11. How many different Ab "isotype" values are in this dataset? 

```{r}
table(abdata$isotype)
```


>Q. How many different "antigen" values are measured? 

```{r}
table(abdata$antigen)
```
>Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```


Let's focus on IgG isotype:  

```{r}
igg <- abdata |> 
          filter (isotype=="IgG")

head(igg)
```

>Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

Make a plot of `MFI-normalised` values for all `antigen` values. 

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() 
```

>Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

PT, PRN, FHA, and FIM2/3 show the biggest changes in IgG levels over time because their boxplots shift more between visits. The other antigens stay mostly low or similar across visits, so they don’t show much change.

The antigen "PT", "FIM2/3" and "FHA" appear to have the widest range of values. 

>Q. Is there a differnce for these responses between aP and wP individuals? 


```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot() 
```

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() +
facet_wrap(~infancy_vac)
```


>Q. Is there a difference with time (i.e. before booster shot vs after booster shot)? 

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot() +
facet_wrap(~visit)
  
```


Lets finish this section by looking at the 2021 dataset IgG PT antigen levels time-course:

```{r}
## filter to 2021 dataset

abdata.21 <- abdata |>
  filter(dataset == "2021_dataset",
         isotype =="IgG", 
         antigen =="PT")

ggplot(abdata.21) +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

```{r}
abdata.21 <- abdata |>
  filter(dataset == "2021_dataset",
         isotype == "IgG",
         antigen == "PT")

ggplot(abdata.21) +
  aes(x = planned_day_relative_to_boost,
      y = MFI_normalised,
      col = infancy_vac,
      group = subject_id) +
  geom_point(alpha = 0.7) +
  geom_line(alpha = 0.5) +

  # thick trend lines
  stat_summary(aes(group = infancy_vac),
               fun = mean,
               geom = "line",
               linewidth = 2.5) +

  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 14, linetype = "dashed") +

  labs(
    title = "CMI-PB 2021 dataset IgG PT",
    subtitle = "Dashed lines at day 0 (pre boost) and day 14 (post boost)",
    x = "Day relative to boost",
    y = "Normalised MFI"
  ) +
  theme_bw()
```





