---
title: "Class 12 Lab"
author: "Sahar Kamali PID:A19156624"
date: "2026-02-12"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Section 1. proportion og G/G in a population 
Downloaded a CSV file from Ensemble < https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39872367-39873367;v=rs12936231;vdb=variation;vf=959765854#373531_tablePanel > 

Here we read this csv file 

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs12936231.csv")
head(mxl)
```


```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```


Now let's look at a different population. I picked the GBR population. 

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs12936231.csv")
head

```
find proportion of G/G 
```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) * 100, 2)
```


This variant that is associated with childhood asthma is more frequent in the GBR population than the MKL population. 

Let's now dig into this further. 

##Section 4: Population Scale Analysis [HOMEWORK]
One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. 

How many samples do we have? 

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

##Sample sizes

```{r}
table(expr$geno)
```


##Median expression values

```{r}
tapply(expr$exp, expr$geno, median, na.rm = TRUE)
```


>Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

The total sample size is 462. The genotype counts are A/A = 108, A/G = 233, and G/G = 121. The median expression values are 31.25 for A/A, 25.06 for A/G, and 20.07 for G/G.



```{r}
library(ggplot2)

```


let's make a boxplot 

```{r}
ggplot(expr) + aes(geno, exp, fill=geno) + 
  geom_boxplot(notch=TRUE)
```
>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

In the boxplot, the A/A group has the highest ORMDL3 expression, A/G is in the middle, and G/G is the lowest. This pattern suggests the G allele is associated with lower ORMDL3 expression, so rs8067378 likely affects ORMDL3 expression.

